---
title: "preop_school_PSM"
author: "Hannu"
date: "2024-04-08"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MatchIt)
library(tidyverse)
source("Z_up.R")
```

## Gastric bypass

### Difference-in-means: TWL, a5_ant_vekt, b_ant_vekt
There are large differences in weight measures between EPEP and SPEP.
The independent variable of interest is preoperative weight school 
(1 = EPEP = Treated ; 0 = SPEP = Control).


```{r GB, echo=FALSE}
d_act_a5_GB |> 
  group_by(o_preop_vektskole) |>
  summarise(TWL = mean(a5_TWL), sd = sd(a5_TWL), se = sd/n() )

with(d_act_a5_GB, t.test(a5_TWL ~ o_preop_vektskole))
with(d_act_a5_GB, t.test(a5_ant_vekt ~ o_preop_vektskole))

with(d_act_a5_GB, t.test(b_ant_vekt ~ o_preop_vektskole))
```
 
 
###  Difference-in-means: pre-treatment covariates
We’ll work with the following covariates for now:
```{r}
# covariates --------------------------------------------------------------
vekt_cov <- c('o_preop_vektskole', 'b_ant_vekt', 'p_alder_v_op', 
              'Female', 'b_beh_diab', 'smoke')
```


Let’s calculate the mean for each covariate by the treatment status:
```{r}
d_act_a5_GB |>
  group_by(o_preop_vektskole) |>
  select(one_of(vekt_cov)) |>
  summarise_all(list(~ mean(., na.rm = T)))
```
The t-test shows that variables: Age and diabetes  
are  are statistically distinguishable between the two groups.
```{r}
lapply(vekt_cov[-1], function(v) {
  t.test(d_act_a5_GB[[v]] ~ d_act_a5_GB[['o_preop_vektskole']])
})
```


## Propensity score estimation
We estimate the propensity score by running a logit model (probit also works) where the outcome variable is a binary variable indicating treatment status. What covariates should you include? For the matching to give you a causal estimate in the end, you need to include any covariate that is related to both the treatment assignment and potential outcomes. 


###
```{r vekttap}
d_act_a5_GB =   d_act_a5_GB |> mutate(a5_WL =   b_ant_vekt - a5_ant_vekt)

m_ps <- glm(formula = o_preop_vektskole ~ a5_WL + p_alder_v_op + Female + 
                      + b_beh_diab + smoke,
            family = binomial(), 
            data = d_act_a5_GB)
summary(m_ps) 
```


 Using this model, we can now calculate the propensity score for each patient. 
 It is simply the patient’s predicted probability of 
 being Treated  (o_preop_vektskole),  given the estimates from the logit model. 
 Below, I calculate this propensity score using 
 predict() and create a tibble that has the propensity score as well as the 
 patient’s actual treatment status.

```{r}
prs_df <- tibble(pr_score = predict(m_ps, type = "response"),
                     o_preop_vektskole = m_ps$model$o_preop_vektskole)
head(prs_df)
```

### Examining the region of common support

After estimating the propensity score, it is useful to plot 
histograms of the estimated propensity scores by treatment status.


```{r}
labs <- paste("Actual school type attended:", c("SPEP", "EPEP"))
prs_df %>%
  mutate(o_preop_vektskole = ifelse(o_preop_vektskole == 1, labs[2], labs[1])) %>%
  ggplot(aes(x = pr_score)) +
  geom_histogram(fill ="lightgreen", color = "black") +
  facet_wrap(~o_preop_vektskole) +
  xlab("Probability of going to EPEP school") +
  theme_minimal()  # 
```
### Executing a matching algorithm
The method we use below is to find pairs of observations that 
have very similar propensity scores, but that differ in their treatment status. 
We use the package MatchIt for this. This package estimates the propensity score 
in the background and then matches observations based on the method of choice 
(“optimal” in this case).

```{r}
vekt_nomiss <- d_act_a5_GB %>%  # MatchIt does not allow missing values
  select(a5_WL, one_of(vekt_cov)) %>%
  na.omit()

mod_match <- matchit(formula =  o_preop_vektskole ~ a5_WL + log(p_alder_v_op) +
                                Female + b_beh_diab + smoke,
                     method = "optimal", 
                     data = vekt_nomiss)
```

## Examining covariate balance in the matched sample

```{r}
summary(mod_match)
```

```{r}
dta_m <- match.data(mod_match)
dim(dta_m)  #[1] 1234  10                     1234 = 617*2

with(dta_m, t.test(a5_WL ~ o_preop_vektskole))
```

Use %TWL instead, gives similar results.
```{r}
TWL_nomiss <- d_act_a5_GB %>%  # MatchIt does not allow missing values
  select(a5_TWL, one_of(vekt_cov)) %>%
  na.omit()

TWL_match <- matchit(formula =  o_preop_vektskole ~ a5_TWL + log(p_alder_v_op) +
                       Female + b_beh_diab + smoke,
                     method = "optimal", 
                     data = TWL_nomiss)

summary(TWL_match)
plot(TWL_match)


dta_m <- match.data(TWL_match)
dim(dta_m)  #[1] 1340   10                     1340 = 670*2

with(dta_m, t.test(a5_TWL ~ o_preop_vektskole))
```


Age distributions?
```{r}
d_act_a5_GB |> group_by(o_preop_vektskole) |> 
  summarise(TWL =   mean(a5_TWL, na.rm = TRUE))

d_act_a5_GB |> 
  ggplot(aes(x=p_alder_v_op, fill=o_preop_vektskole, group = o_preop_vektskole)) +
  geom_density(alpha = 0.4)
```



```{r}
ps_var =  function(dta= dta_m, variable= "p_alder_v_op"){
support <- c(min(dta[[variable]]), max(dta[[variable]]))

ggplot(dta, aes(x = distance, y = p_alder_v_op, 
                 group = o_preop_vektskole, color = o_preop_vektskole)) +
  geom_point(alpha = 0.2, size = 1.3) +
  geom_smooth(method = "loess", se = F) +
  xlab("Propensity score") +
  ylab(variable) +
  theme_bw() + theme(legend.position = "none") +
  ylim(support)
}

ps_var(dta_m, variable =  "p_alder_v_op")
ps_var(dta_m, variable =  "a5_TWL")
```

 